# *clr-lasso* {#clr}

Penalised regression is a powerful approach for variable selection in high dimensional settings [@zou2005regularization; @tibshirani1996regression; @le1992ridge]. It can be adapted to compositional data analysis (CoDA) by previously transforming the compositional data with the centered log-ratio transformation (clr). 

*Clr-lasso* implements penalised regression using the R package *glmnet* after clr transformation. Clr transformation corresponds to centering the log-transformed values: 

$$clr(x) = clr(x_{1},...,x_{k}) = (log(x_{1})-M,...,log(x_{k})-M)$$

Where $i=1,...,k$ microbial variables, $x_{k}$ is the counts of variable $k$, $M$ is the arithmetic mean of the log-transformed values.

$$M = \frac{1}{k}\sum_{i=1}^{k}log(x_{i})$$


We also generated a wrapper function called *glmnet_wrapper()* that will help us to handle the output of *glmnet*. The *glmnet_wrapper()* function is uploaded via **functions.R**.


## Crohn case study

First we perform the clr transformation of the abundance table **x_Crohn** as follows (the log-transformation requires a matrix of positive values and thus the zeros have been previously added with an offset of 1):  


```{r}
z_Crohn <- log(x_Crohn)
clrx_Crohn <- apply(z_Crohn, 2, function(x) x - rowMeans(z_Crohn))
```

We implement penalised regression with function *glmnet()*. This function requires the outcome **Y** to be numeric. The file "Crohn_data.RData" contains **y_Crohn_numeric**, a numerical vector of disease status with values 1 (CD) and 0 (not CD). 

Penalised regression requires the specification of the penalization parameter $\lambda$. The larger the value of $\lambda$, the less variables will be selected. In previous analysis of this dataset with *selbal*, a balance with 12 variables was determined optimal to discriminate the CD status [@rivera2018balances]. For ease of comparison, we will specify the penalisation parameter that results in the selection of 12 variables for *clr-lasso*. 

To identify the value of $\lambda$ that corresponds to 12 variables selected we implement *glmnet()* function on the clr transformed values and with the specification that the response family is **binomial**:


```{r}
Crohn.test_clrlasso <- glmnet(x = clrx_Crohn, y = y_Crohn_numeric, 
                              family = 'binomial', nlambda = 30)
```

The output of *glmnet()* can be visualised with the selection plot:

```{r CDlasso, fig.cap = 'Selection plot of Crohn data'}
plot(Crohn.test_clrlasso, xvar = 'lambda', label = T)
```

In Figure \@ref(fig:CDlasso), each curve corresponds to a variable (i.e. genus). It shows the path of its coefficient against different $log(\lambda)$ values. At each $log(\lambda)$, the shown curves indicate the number of nonzero coefficients. In the plot command, if **label = T**, each curve will be annotated with variable index. 

The numerical output of *glmnet()* will help us to select the value of $\lambda$. It provides the number of selected variables or degrees of freedom of the model (Df), the proportion of explained deviance for a sequence of values of $\lambda$:

```{r}
Crohn.test_clrlasso
```

We can see that the value of $\lambda$ that will select 12 variables is between 0.037 and 0.051. We perform again *glmnet()* with the specification of a finer sequence of $\lambda$ (between 0.03 and 0.05 and an increment of 0.001): 

```{r }
Crohn.test2_clrlasso <- glmnet(x = clrx_Crohn, y = y_Crohn_numeric, 
                               family = 'binomial', lambda = seq(0.03, 0.05, 0.001))
Crohn.test2_clrlasso
```

According to this result, we select $\lambda = 0.043$ and implement *glmnet()*:

```{r}
clrlasso_Crohn <- glmnet(x = clrx_Crohn, y = y_Crohn_numeric, 
                         family = 'binomial', lambda = 0.043)
```


The same as *selbal*, we use function *glmnet_wrapper()* to handle the results of *clr-lasso* with the output from *glmnet()*:


```{r}
Crohn.results_clrlasso <- glmnet_wrapper(clrlasso_Crohn, X = clrx_Crohn)
```

We can get the number of selected variables:

```{r}
Crohn.results_clrlasso$numVarSelect
```

and the names of the selected variables:

```{r}
Crohn.results_clrlasso$varSelect
```


## HFHS-Day1 case study

The analysis on HFHSday1 data is similar to Crohn data. We first perform the clr transformation:

```{r}
z_HFHSday1 <- log(x_HFHSday1)
clrx_HFHSday1 <- apply(z_HFHSday1, 2, function(x) x-rowMeans(z_HFHSday1))
```

The outcome **y_HFHSday1** is converted to a numeric vector.

```{r}
y_HFHSday1_numeric <- as.numeric(y_HFHSday1)
```

We implement *glmnet()* and visualise the results (Figure \@ref(fig:HFHSlasso)). 

```{r}
HFHS.test_clrlasso <- glmnet(x = clrx_HFHSday1, y = y_HFHSday1_numeric, 
                             family = 'binomial', nlambda = 30)
```

```{r HFHSlasso, fig.cap = 'Lasso plot of HFHSday1 data'}
plot(HFHS.test_clrlasso, xvar = 'lambda', label = T)
```

The explanation of Figure \@ref(fig:HFHSlasso) is the same as Figure \@ref(fig:CDlasso).

The numerical output of *glmnet()* will help us to decide the penalty term $\lambda$:

```{r}
HFHS.test_clrlasso
```

For comparison purposes we set the penalty term $\lambda$ equal to 0.03 (**lambda = 0.03**) that results in the selection of 10 OTUs. 

```{r}
clrlasso_HFHSday1 <- glmnet(x = clrx_HFHSday1, y = y_HFHSday1_numeric, 
                         family = 'binomial', lambda = 0.03)
```

We then use function *glmnet_wrapper()* to handle the results:

```{r}
HFHS.results_clrlasso <- glmnet_wrapper(result = clrlasso_HFHSday1, X = clrx_HFHSday1) 
```

Then we get the number of selected variables:

```{r}
HFHS.results_clrlasso$numVarSelect

```

and the names of selected variables:

```{r}
HFHS.results_clrlasso$varSelect
```

We also extract the taxonomic information of these selected OTUs.

```{r}
HFHS.tax_clrlasso <- taxonomy_HFHS[which(rownames(taxonomy_HFHS) %in% 
                                           HFHS.results_clrlasso$varSelect), ]
kable(HFHS.tax_clrlasso[ ,2:6], booktabs = T)
```
